---
title: Metadata checks
eval: false
---

Check metadata for projects we have control over. 
If we find any problems we have the **power** (`r emo::ji("strong")`) to fix them (`r emo::ji("grin")`).





## Check tag details

```{r}
tag_deps <- map(dbs, \(x) tbl(x, "tagDeps") |> collect()) |>
  list_rbind() |>
  mutate(tsStart = as_datetime(tsStart), tsEnd = as_datetime(tsEnd))

tag_deps |>
  select(projectID, tagID, deployID, tsStart, tsEnd, speciesID) |>
  summary()
```


TODO: Questions

- Can the end be after the current date? e.g., 2026-12-16...
- Should there be missing start/end dates?

```{r}
select(tag_deps, projectID, tagID, deployID, tsStart, tsEnd, speciesID) |>
  filter(is.na(tsStart))
```


### Check for multiple tag deployments
```{r}
multi_tags <- map(dbs, \(x) tbl(x, "tagDeps") |> 
                    select(tagID, deployID, projectID, tsStart, tsEnd) |>
                    mutate(n = n(), .by = "tagID") |>
                    filter(n > 1) |>
                    collect()) |>
  list_rbind() |>
  mutate(tsStart = as_datetime(tsStart), tsEnd = as_datetime(tsEnd))
```

Multiple deployments shouldn't be able to overlap... right?

```{r}
multi_tags |>
  group_by(tagID) |>
  filter(tsEnd >= lead(tsStart) | tsStart <= lag(tsEnd))
```

Triple deploy ids all at the same time???
```{r}
tbl(dbs[["551"]], "allruns") |>
  select(motusTagID, tagDeployID, tsBegin) |>
  filter(motusTagID == 68470) |>
  collect() |>
  mutate(tsBegin = as_datetime(tsBegin))
```


### Triple deploy ids all at the same time???
```{r}
tbl(dbs[["168"]], "allruns") |>
  select(motusTagID, tagDeployID, tsBegin) |>
  filter(motusTagID == 26448) |>
  collect() |>
  mutate(tsBegin = as_datetime(tsBegin)) |>
  filter(tsBegin > "2017-08-01 20:00:00")
```
