---
title: "Citations"
---

This section contains citations that we should remember to include.

### eBird Status and Trends

In addition to citing the `ebirdst` package (below). 
We *also* have to cite the eBird data itself.

eBird data requires attributions, citations, and disclaimers

For the most up-to-date instructions see Part 2:  <https://science.ebird.org/en/status-and-trends/products-access-terms-of-use>



### Packages

You probably only need to cite **motus**, **ebirdst**, and possibly the **tidyverse**, 
but you should keep track of the packages used and their versions 
(see General Info, note that if an workflow step was updated separately, see
the most up-to-date package versions under that pages "General Info").

Unfortuntaly journals do not often want you to cite general data manipulation 
and visualization packages (which is most of what we use).

```{r}
#| code-fold: true
#| echo: false
#| results: asis
cat("#### motus\n")
c <- citation("motus") 
print(c, "text")

cat("\n#### ebirdst (for access to range maps)\n")
library(ebirdst)

cat("\n#### tidyverse (for general data management and figures)\n")
citation("tidyverse") |> print(style = "text")
```

### Maps

Note that map tiles should be attributed. Best is to include that attribution
 in the figure itself, but you can also include it in the caption.
 
We use [CARTO light tiles](https://carto.com/basemaps).
 

### General Info
```{r}
devtools::session_info()
```