---
title: Setup
---

```{r}
#| include: false
knitr::read_chunk("XX_setup.R")
```

This page explains the setup code defined in `XX_setup.R` which is sourced 
at the start of each workflow step.

## Packages

```{r packages}
#| message: false
```

## Motus credentials 

Here we set the credentials for this session 
(so you don't have to manually type it in every time)
 
This code expects that you store your Motus login and password as 
`URBAN_USER` and `URBAN_PASSWORD` in an .Renviron file

- learn more about this file: https://rstats.wtf/r-startup.html#renviron
- easily edit this file with `usethis::edit_r_environ()`

```{r credentials}
```

## Define projects 

Here we're using a small set of pilot projects

- Urban Ecology of Birds in the Lower mainland of BC #484
- Montreal Urban Forests #551 (191 tags but 62 on resident cardinals)
- Montreal Connectivity #373 (105 tags)
- Montreal moult migrant research #168 (113 tags)

Additionally we're using open motus projects selected via 
[Selecting Projects](01_select_projects.qmd).

```{r projects}
```

## Define contants

Here we define any cutoffs or constants we will be using

- `max_flight_speed` is the maximum speed of travel we think is possible (with a buffer)
- `max_tower_dist` is the maximum distance between towers/receivers within which
a bird could be detected at both towers (and thus flight speed has no bearing)
- `bout_cutoff` is the maximum amount of time between two runs at a specific 
station before those runs will be considered separate detection bouts.

```{r constants}
```


## Data

Get species lists from naturecounts taxonomy and join this with the species
metadata from the motus data bases. 

We only need to do this to one database, as the `metadata()` data step in [Download/Update data](02_download.html) adds full species lists to all databases (so all database metadata are the same).

### Species
```{r data_sp}
```

### Databases

#### Load existing databases
```{r db_load}
```

#### Establish included species
- We match existing species in the databases to the species list from naturecounts
- We rename scientific names to omit subspecies (e.g., myrtle yellow-rumped warbler)
  [required for later eBird range steps]
  
```{r db_species}
```


## Functions
These functions are sourced from `XX_functions.R`
```{r}
#| file: XX_functions.R
```

