---
title: Setup
---

```{r}
#| include: false
knitr::read_chunk("XX_setup.R")
```

This page explains the setup code defined in `XX_setup.R` which is sourced 
at the start of each workflow step.

## Packages

```{r packages}
#| message: false
```

## Motus credentials 

Here we set the credentials for this session 
(so you don't have to manually type it in every time)
 
This code expects that you store your Motus login and password as 
`URBAN_USER` and `URBAN_PASSWORD` in an .Renviron file

- learn more about this file: https://rstats.wtf/r-startup.html#renviron
- easily edit this file with `usethis::edit_r_environ()`

```{r credentials}
```

## Define projects 

Here we're using a small set of pilot projects

- Urban Ecology of Birds in the Lower mainland of BC #484
- Montreal Urban Forests #551 (191 tags but 62 on resident cardinals)
- Montreal Connectivity #373 (105 tags)
- Montreal moult migrant research #168 (113 tags)

```{r projects}
```

## Define data to include

TODO: What are 'near-passerines'? cf https://en.wikipedia.org/wiki/Near_passerine

```{r data}
```

## Functions
These functions are sourced from `XX_functions.R`
```{r}
#| file: XX_functions.R
```

